---
output:
  html_document: default
  pdf_document: default
---
White Wine Quality by Irina Ford
========================================================

**White Wine Quality** dataset was selected for this project. The data contain
**11** chemical properties for **4898** samples of wine including _acidity,
residual sugar, sulfur dioxide, chlorides, sulphates, density, pH,_ and
_alcohol_. The wine _quality_ attribute (median of several evaluations made by
wine experts) is also included. The goal was to explore which chemical
properties affect the quality of white wine the most.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(gridExtra)
library(GGally)
library(dplyr)

# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

```

The data set was loaded into RStudio with the summary of attributes as follows: 

```{r echo=FALSE, Load_the_Data}
# Load csv file, and look at the variables types and statistics
wine <- read.csv('C:/Users/Irina/Documents/MY_PROJECTS/_WGU/_C751/Project_White_Wine_Quality/wineQualityWhites.csv')
str(wine)
summary(wine)

```

# Univariate Plots Section

Now let's look at distribution plots of chemical properties.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
# histograms - fixed acidity, volatile acidity, citric acid
p1 <- ggplot(aes(fixed.acidity), data = wine) +
    geom_histogram()
p2 <- ggplot(aes(volatile.acidity), data = wine) +
    geom_histogram()
p3 <- ggplot(aes(citric.acid), data = wine) +
    geom_histogram()

grid.arrange(p1, p2, p3, ncol = 3)
```

Fixed and volatile acidity and citric acid seem to be normally distributed with
some outliers on the high end of scale


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2}
# histograms - residual sugar, log10(chlorides)
p4 <- ggplot(aes(residual.sugar), data = wine) +
    geom_histogram()
p5 <- ggplot(aes(log10(chlorides)), data = wine) +
    geom_histogram()
grid.arrange(p4, p5, ncol = 2)

```

Both residual sugar and chlorides distribution are slightly skewed to the right.
I used log10 scale to plot chlorides.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots3}
# histograms - free sulfur dioxide, total sulfur dioxide
p6 <- ggplot(aes(free.sulfur.dioxide), data = wine) +
    geom_histogram()
p7 <- ggplot(aes(total.sulfur.dioxide), data = wine) +
    geom_histogram()
grid.arrange(p6, p7, ncol = 2)

```

Free and total sulfur dioxide looks normally distributed with some outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4}
# histograms - density, pH
p8 <- ggplot(aes(density), data = wine) +
    geom_histogram()
p9 <- ggplot(aes(pH), data = wine) +
    geom_histogram()
grid.arrange(p8, p9, ncol = 2)
```

Density has outliers, but pH Histogram is normally distributed and does not
seem to have outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots5}
# histograms - sulphates, alcohol
p10 <- ggplot(aes(sulphates), data = wine) +
    geom_histogram()
p11 <- ggplot(aes(alcohol), data = wine) +
    geom_histogram()
grid.arrange(p10, p11, ncol = 2)
```

Sulphates and alcolol are both very slightly skewed to the right.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots6}
# Histogram - quality
ggplot(aes(quality), data = wine) +
  geom_histogram(stat="count") +
  scale_x_continuous(breaks = seq(1,10,1))
```

```{r echo=FALSE, quality}
table(wine$quality)
```

**Quality** attribute has discrete values in range from **3** to **9** and can
also be considered as ordered factor variable.
Looking at the distribution and table of counts for quality there are much fewer
samples for very poor and very good quality wines with the medium quality values
of **5** and **6** being dominant in the date set. We can create quality
buckets by combining some values together in order to increase statistics for
end points quality and find correlation with other chemical properties.



```{r echo = FALSE, quality_buckets}
# combining quality values 3 + 4 and 8 + 9 to make 5 buckets
wine$quality.bucket <- cut(wine$quality, c(2,4,5,6,7,9))
table(wine$quality.bucket)

```


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots7}
# histogram - quality buckets 
ggplot(aes(quality.bucket), data = wine) + geom_histogram(stat='count')
```

# Univariate Analysis
### What is the structure of your dataset?
There are 4898 white wine samples with 11 chemical properties measured: fixed
acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur
dioxide, total sulfur dioxide, density, pH, sulphates, alcohol. There is also
wine quality variable, which is a score that can take values from 0 (very bad)
to 10 (very excellent), it can be used as ordered factor variable in our
data set.

Observations:
* wine quality changes in range from 3 to 9. 75% of all samples have
medium quality of 5-6.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in this dataset is wine quality and how it changes
based on changes in chemical properties.

### What other features in the dataset do you think will help support your
### investigation into your feature(s) of interest?
Alcohol content was found to be the main factor affecting wine quality, but some
other properties like density, total sulfur dioxide, volatile acidity and
chlorides also could potentially affect wine quality.   

### Did you create any new variables from existing variables in the dataset?
I created a new factor variable quality bucket to combine some quality groups
together to improve end points statistics.  

### Of the features you investigated, were there any unusual distributions?
### Did you perform any operations on the data to tidy, adjust, or change
### the form of the data? If so, why did you do this?
Most of the distributions of chemical properties were slightly positively skewed
with some outliers on the high end. I used the log10 scale to show distribution
of chlorides content.


# Bivariate Plots Section

For a quick check of possible correlations between attributes I used ggpairs
plot. Noticeble correlation exist between:
* density and alcohol and residual sugar (this corresponds with the description
of variables in _wineQualityInfo.txt_ file).

Moderate correlation exist between:
* alcohol and residual sugar, chlorides and total sulfur dioxide.
* pH and fixed acidity
* density and total sulfur dioxide
* free sulfur dioxide and total sulfur dioxide (those should correlate as the
former is a part of the latter).


```{r echo = FALSE, message=FALSE, warning=FALSE, QC, fig.width=9,fig.height=7}
# create quick look plot to check possible relationships between
# variables in data set

wine$X <- NULL
ggpairs(wine,
     lower = list(continuous = wrap("points", shape = I('.'))),
     upper = list(combo = wrap("box", outlier.shape = I('.')))) +
  theme(text = element_text(size=10), 
     axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size=7),
     panel.grid.minor = element_blank(), 
     panel.grid.major = element_blank())  # rotate and resize axis labels
                                          # and remove grid lines

```

I can also use quality as numeric variable to find possible variables
that affect it:

```{r echo=FALSE, Bivariate_Plots}
# calculate correlation coefficients for quality with other attributes
cor_quality <- subset(wine, select = -c(quality.bucket))
cor(x = cor_quality, y = wine$quality)
```
Alcohol has highest influence on quality, followed by density and chlorides. 
Let's look at some of the relationships.
Here I plot density vs alcohol and residual sugar with outliers removed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots2}
# scatter plot of alcohol vs density
ggplot(aes(x = alcohol, y = density), data = wine) +
 ylim(quantile(wine$density, 0.0), quantile(wine$density, 0.99)) +
 geom_point() +
 geom_smooth(method='lm', color='red')
 
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots3}
# scatter plot of residual sugar vs density
ggplot(aes(x = residual.sugar, y = density), data = wine) +
 geom_point() +
 xlim(quantile(wine$residual.sugar, 0.0), quantile(wine$residual.sugar, 0.99)) +
 ylim(quantile(wine$density, 0.0), quantile(wine$density, 0.99)) +
 geom_smooth(method='lm', color='red')

```

I can look at the boxplots of different variables as they change with quality
as in the plots below. Here we can possibly infer that quality increases with
increase in alcohol content.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots4}
# box plots of alcohol content for each quality value 
qplot(x = as.factor(quality), y=alcohol,
      data = wine, geom = 'boxplot') 

```

Most of the boxplots do not reveal any effect on quality, so I am going to try
to use quality buckets that I created earlier to find mean values of chemical
properties for each bucket and plot them.
After looking at all the attributes the most interesting are as follows:

**Median chlorides value vs quality**

This scatter plot has mean values of chlorides content superimposed for
different quality buckets and shows that decrease in chlorides content leads to
quality improvement.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots5}

# scatter plot  - quality and mean chlorides
ggplot(data = wine, aes(x=quality.bucket, y=chlorides)) +
  geom_jitter(width=0.35, height=0, color="darkgrey") +
  geom_point(stat = "summary", fun.y = "mean", shape = 24, size = 4,
             fill = "blue")

```


**Mean ratio of total to free sulfur dioxide vs quality**

Graph shows increase in quality with decreasing of the ratio of total to
free sulfur dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots6}
# scatter plot  - quality and mean total sulfur dioxide/free sulfur dioxide
ggplot(data = wine,
       aes(x=quality.bucket, y=total.sulfur.dioxide/free.sulfur.dioxide)) +
 geom_jitter(width=0.35, height=0, color="darkgrey") +
 geom_point(stat = "summary", fun.y = "mean", shape = 24, size = 4,
       fill = "blue")

```

**Mean alcohol content vs quality**

Graph shows increase in quality with icrease in alcohol content as we already
have seen before.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots7}
# scatter plot  - quality and mean alcohol
ggplot(data = wine, aes(x=quality.bucket, y=alcohol)) +
  geom_jitter(width=0.35, height=0, color="darkgrey") +
  geom_point(stat = "summary", fun.y = "mean", shape = 24, size = 4,
            aes(fill = "Mean")) +
  scale_fill_manual("", values = c("Mean" = "blue"))

```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the
### investigation. How did the feature(s) of interest vary with other features
### in the dataset?
Using quality as numerical variable the strongest correlation was observed with
alcohol content (R^2 = 0.436), followed by density (R^2 = -0.307). Increase in
alcohol contend and decrease in density both caused wine quality to increase.

### Did you observe any interesting relationships between the other features
### (not the main feature(s) of interest)? What was the strongest relationship
### you found?
The strongest correlations between chemical properties themselves were observed
between density and residual sigar (R^2 = 0.839) and density and alcohol
(R^2 = -0.78), this corresponds with the description of variables
in _wineQualityInfo.txt_ file). The decrease in chlorides content and
in the ratio of total to free sulfur dioxide also seem to improve vine quality.

# Multivariate Plots Section

The graphs in the earlier section can be enhanced by adding the third variable
_quality_ as color. The plot below shows that the quality of wine is mostly
higher for lower values of density and higher values of alcohol content. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
# create multivariate plot of density-alcohol-quality

ggplot(aes(x = alcohol, y = density, colour = as.factor(quality.bucket)),
       data = wine) + geom_point() +
  theme(panel.background = element_rect(fill='darkgrey'),
       legend.key=element_rect(fill='darkgrey')) +
  scale_color_brewer(palette = "RdYlBu",
       guide = guide_legend(title = 'Quality', reverse = T,
       override.aes = list(alpha = 1, size = 2))) +
  ylim(quantile(wine$density, 0.0), quantile(wine$density, 0.99)) 
```

The graph shows separation of values driven by quality of wine. Looks like
higher quality samples lay in the area of lower densities and slightly lower
residual sugar values. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2}
# create multivariate plot of density-residual sugar-quality
ggplot(aes(x = residual.sugar, y = density, colour = as.factor(quality.bucket)),
       data = wine) + geom_point() +
  theme(panel.background = element_rect(fill='darkgrey'),
       legend.key=element_rect(fill='darkgrey')) +
  scale_color_brewer(palette = "RdYlBu",
       guide = guide_legend(title = 'Quality', reverse = T,
       override.aes = list(alpha = 1, size = 2))) +
  ylim(quantile(wine$density, 0.0), quantile(wine$density, 0.99)) +
  xlim(quantile(wine$residual.sugar, 0.0), quantile(wine$residual.sugar, 0.99))
```

On this graph the higher quality samples have higher alcohol content and
slightly lower chlorides level. Although the relationship is subtle. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots3}
# create multivariate plot of alcohol-chlorides-quality
ggplot(aes(x = log10(chlorides), y = alcohol,
        colour = as.factor(quality.bucket)), data = wine) +
  geom_point() +
  theme(panel.background = element_rect(fill='darkgrey'),
        legend.key=element_rect(fill='darkgrey')) +
  scale_color_brewer(palette = "RdYlBu",
        guide = guide_legend(title = 'Quality', reverse = T,
        override.aes = list(alpha = 1, size = 2)))
```

The plot below is arguable as well, but potentially total to free sulfur
dioxide ratio could be influencing wine quality. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots4}
# create multivariate plot of total sulfur dioxide/free sulfur dioxide - 
# alcohol-quality
ggplot(aes(x = alcohol, y = log10(total.sulfur.dioxide/free.sulfur.dioxide),
        colour = as.factor(quality.bucket)), data = wine) +
  geom_point() + 
  theme(panel.background = element_rect(fill='darkgrey'),
        legend.key=element_rect(fill='darkgrey')) +
  scale_color_brewer(palette = "RdYlBu",
        guide = guide_legend(title = 'Quality', reverse = T,
        override.aes = list(alpha = 1, size = 2)))
```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the
### investigation. Were there features that strengthened each other in terms
### of looking at your feature(s) of interest? 
There appears to be an improvement in wine quality with decreasing density and
increasing alcohol content. Decrease in chlorides content also positevely
affects quality. There arguably might be influence of residual sugar and total
sulfur dioxide as well.   

### Were there any interesting or surprising interactions between features?
I did not find any correlation with volatile acidity although in description of
attributes it was mentioned as potential influencer on wine quality. 

------


# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# histogram - wine quality 
  hist(wine$quality, main = 'Wine Quality Distribution',
       xlab = "Quality",
       xlim = c(0, 10),
       col="lightgreen",
       ylab = "Number of Samples", breaks = c(0:10))
```

### Description One

Distribution of wine quality ratings by wine experts. Most of samples have
medium scores of 5 and 6 with very few observation of poor and excellent
quality wines.  

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# boxplots of alcohol content for each quality category
ggplot(aes(as.factor(quality), alcohol), data = wine) +
  geom_boxplot() +
  labs(title = "Alcohol Content by Quality", x = "Quality", y = "Alcohol")

```

### Description Two

Highest quality samples tend to have an elevated levels of alcohol content
compared with average and poor quality samples.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# scatter plot alcohol vs chlorides with quality added as color
ggplot(aes(x = log10(chlorides), y = alcohol,
        colour = as.factor(quality.bucket)), data = wine) +
  geom_point() +
  theme(panel.background = element_rect(fill='darkgrey'),
        legend.key=element_rect(fill='darkgrey')) +
  scale_color_brewer(palette = "RdYlBu",
        guide = guide_legend(title = 'Quality', reverse = T,
        override.aes = list(alpha = 1, size = 2))) + 
  ggtitle('Alcohol Content by Chlorides (log10) and Quality') +
  labs(y = "Alcohol Content (% by Vol)", x="Chlorides Content (g/dm3)")

```

### Description Three

Both increase in alcohol content and decrease in chlorides content have
positive influence on wine quality as shown in this multivariate plot.

------

# Reflection

Exploratory data analysis was performed to determine the main influencing
factors on the quality of white wine. Quality was found to be affected by
alcohol, chlorides content, total and free sulfur dioxide, and density.

Some correlations are subtle and might be a subject for further investigation.
Classification or regression models could be build to predict the quality
of wine based on its measured chemical properties.

In regard with my personal experience with R, I started working on the project
by loading the data set and tried to understand if there were any possible
relashionships between quality as a factor variable and other properties,
so I started with quick look plot and then tried to find
what visualizations could highlight those relationships the best.
I found difficult to fine tune some plots (such as adding background, axes 
labels, specific palettes, etc.). I also had problem with file permissions and
packages download (especially knitr).

Overall, I found this project very useful and interesting as R provides
enormous possibilities for exploring, visualizing, and modeling data and is
fairly easy to understand.